---
title: "Convert 10x genomics chromatin accessibility matrix to ranged summarized experiment"
output:
  rmarkdown::html_document:
    highlight: pygments
    toc: true
---

# Introduction

In this example we use the following data set: 

500 1:1 mixture of fresh frozen human (GM12878) and mouse (A20) cells

Single Cell ATAC Dataset by Cell Ranger ATAC 1.0.1

1:1 mixture of human (GM12878) and mouse (A20) cells.This is a classic human-mouse mixture experiment to demonstrate single cell behavior.

- ~750 transposed nuclei were loaded.
- 481 human and 352 mouse nuclei were recovered.
- Sequenced on Illumina NovaSeq with approximately 24k read pairs per cell.
- 50bp read1, 8bp i7 (sample index), 16bp i5 (10x Barcode), 49bp read2.
- Published on December 17, 2018

This dataset is licensed under the Creative Commons Attribution license.

https://support.10xgenomics.com/single-cell-atac/datasets/1.0.1/atac_v1_hgmm_500

10x uses Cell Ranger ATAC 1.0.1 for bioinformatic processing and produces a peak by cell matrix. 


# Download Data

Download the file corresponding to Peak by cell matrix (filtered)	148.77 MB	87faab53ab541ab5f5d5af3b9748917b 

We download the peak by cell matrix via the terminal using curl, although other options are available on thier webpage. 

```{bash eval=F}
cd localDir10xData
curl -O http://cf.10xgenomics.com/samples/cell-atac/1.0.1/atac_v1_pbmc_10k/atac_v1_pbmc_10k_filtered_peak_bc_matrix.tar.gz
```

Unpack the tar.gz download via the terminal (method may depend on operating system)

```{bash eval=F}
tar -xzf atac_v1_pbmc_10k_filtered_peak_bc_matrix.tar.gz
```

The resulting folder contains 3 files:

- matrix.mtx:  peak by cell chromatin accessiblilty matrix in MatrixMarket format.  We can use R package "Matrix" to read straight to a sparse matrix
- peaks.bed bed file:  corresponding to rows of matrix.mtx
- barcodes.tsv:  barcodes corresponding to columns of matrix.mtx


# Create ranged summarized experiment from 10x peak by cell matrix 

set data10xDir to appropriate local folder
```{r}
data10xDir = "~/Dropbox/Documents/statGen/scATACseq/10xgenomics/atac_v1_hgmm_500/filtered_peak_bc_matrix"
```

install destin R package
```{r  message=FALSE}
yourPathToDestinRepo = "~/Documents/gitRepos/destin"
install.packages(file.path(yourPathToDestinRepo,"package"), repos = NULL, type = "source")
```

load Destin
```{r  message=FALSE}
suppressMessages( library(destin, quietly = T) )
```

Create ranged summarized experiment from 10x peak by cell matrix 
```{r}
rse = createRSEfrom10xMatrix(data10xDir)
```

here is a description of the ranged summarized experiment  

column data is a description of the cells
```{r}
colData(rse)
```

rowRanges is a description of the peaks
```{r}
rowRanges(rse)
```

assay is the binary chromatin accessibility matrix in sparse matrix format

- rows are peaks 
- columns are cells

```{r}
assay(rse)[1:10,1:10]
```

