---
title: "Convert 10x genomics chromatin accessibility matrix to ranged summarized experiment"
output:
  rmarkdown::html_document:
    highlight: pygments
    toc: true
---

# Introduction

In this example we use the following data set: 

5k Peripheral blood mononuclear cells (PBMCs) from a healthy donor

Single Cell ATAC Dataset by Cell Ranger ATAC 1.0.1

Peripheral blood mononuclear cells (PBMCs) from a healthy donor.

- ~6700 transposed nuclei were loaded.
- 5335 nuclei were recovered.
- Sequenced on Illumina NovaSeq with approximately 42k read pairs per cell.
- 50bp read1, 8bp i7 (sample index), 16bp i5 (10x Barcode), 49bp read2.
- Published on December 17, 2018

This dataset is licensed under the Creative Commons Attribution license.

https://support.10xgenomics.com/single-cell-atac/datasets/1.0.1/atac_v1_pbmc_5k

10x uses Cell Ranger ATAC 1.0.1 for bioinformatic processing and produces a peak by cell matrix. 

Approximate time to run (8 core macbook pro): ~ 13 min

```{r}
startTime = Sys.time()
```

# Download Data

Download the file corresponding to Peak by cell matrix (filtered)	97.86 MB	03ea4420a7bdaadd8c1cab8d4a0cf2e9

We download the peak by cell matrix via the terminal using curl, although other options are available on thier webpage. 

```{bash eval=F}
cd localDir10xData
curl -O http://cf.10xgenomics.com/samples/cell-atac/1.0.1/atac_v1_pbmc_5k/atac_v1_pbmc_5k_filtered_peak_bc_matrix.tar.gz
```

Unpack the tar.gz download via the terminal (method may depend on operating system)

```{bash eval=F}
tar -xzf atac_v1_pbmc_5k_filtered_peak_bc_matrix.tar.gz
```

The resulting folder contains 3 files:

- matrix.mtx:  peak by cell chromatin accessiblilty matrix in MatrixMarket format.  We can use R package "Matrix" to read straight to a sparse matrix
- peaks.bed bed file:  corresponding to rows of matrix.mtx
- barcodes.tsv:  barcodes corresponding to columns of matrix.mtx

set data10xDir to appropriate local folder
```{r}
data10xDir = "~/Dropbox/Documents/statGen/scATACseq/10xgenomics/atac_v1_pbmc_5k/filtered_peak_bc_matrix"
```

install destin R package
```{r  message=FALSE}
yourPathToDestinRepo = "~/Documents/gitRepos/destin"
install.packages(file.path(yourPathToDestinRepo,"package"), repos = NULL, type = "source")
```

load Destin
```{r  message=FALSE}
library(destin, quietly = T) 
```

# Create ranged summarized experiment from 10x peak by cell matrix 
```{r}
rse = createRSEfrom10xMatrix(data10xDir)
```


# test memory limits
```{r}
countMat = assay(rse)
countMatBig = cbind(countMat, countMat, countMat, countMat, countMat) 
countMatBigger = cbind(countMatBig, countMatBig, countMatBig)
pca = irlba(t(countMatBigger), nv = 2)
```


here is a description of the ranged summarized experiment  

column data is a description of the cells
```{r}
colData(rse)
```

rowRanges is a description of the peaks
```{r}
rowRanges(rse)
```

assay is the binary chromatin accessibility matrix in sparse matrix format

- rows are peaks 
- columns are cells

```{r}
assay(rse)[1:10,1:10]
```

# Annotate 
```{r}
  model = "hg19"
  rse = annotateRSE(rse, model)
```

# QC
```{r}
  rse = doQC(rse, regionSumCutoff = 5, cellSumCutoffSDs = 3)
```

# Estimate number of clusters 
```{r}
clusterEst = estimateNClusters(rse, nClustersRange = 2:20)
nClusters = clusterEst$nClustersList$logLikeElbow
plotNClusters(clusterEst)
```

# Perform Destin clustering 
```{r}
#set nCores to an appropriate value based on your system
nCores = 7 
clusterResults = destinGrid (rse, nClusters = nClusters, nCores = nCores)
clusterResults$cluster
```

# Plot results
```{r}
plotCluster(clusterResults)
```

Time to run
```{r}
TotalTime = Sys.time() - startTime
TotalTime
```

